---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r load activity.csv}
## URL of the dataset
sourceDataURL <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip'
## File names for data
fileList <- c('activity.csv')
## Temporary File for storing the downloaded zip then extract the files we want into R
temp <- tempfile()
download.file(sourceDataURL, temp, method = "libcurl")
## Read the file from the source
activity.data <- read.csv("activity.csv", colClasses = c("numeric", "character", "numeric"))
## Make real dates
activity.data[, "date"] <- as.Date(activity.data[, "date"], "%Y-%m-%d")
## Summarize data
## Total steps by day
total.daily.steps <- aggregate(activity.data[, "steps"], by = list(activity.data[, "date"]), FUN = sum, na.rm = TRUE)
colnames(total.daily.steps) <- c("date", "daily.steps")
## Mean steps by interval
total.interval.steps <- aggregate(activity.data[, "steps"], by = list(activity.data[, "interval"]), FUN = mean, na.rm = TRUE)
colnames(total.interval.steps) <- c("interval", "mean.steps")
```
### Summary of the data read in
```{r summary of data}
summary(activity.data)
```

## What is mean total number of steps taken per day?
### Histogram of the number of steps daily
```{r mean steps per day histogram}
hist( 
      total.daily.steps[, "daily.steps"]
     ,xlab = "Steps Per Day"
     ,ylab = "Frequency"
     ,breaks = 10
     ,main = "Histogram of Total Daily Steps (na removed)"
    )
```

### Mean and Median daily steps
```{r mean and median}
mean(total.daily.steps[, "daily.steps"])
median(total.daily.steps[, "daily.steps"])
```

The mean is `r mean(total.daily.steps[, "daily.steps"])` and the median is `r median(total.daily.steps[, "daily.steps"])`

## What is the average daily activity pattern?

Time series plot of interval activity
```{r time series plot}
plot(
      total.interval.steps[, "interval"]
     ,total.interval.steps[, "mean.steps"]
     ,type = "l"
     ,xlab = "Interval"
     ,ylab = "Average Number of Steps"
     ,main = "Average Daily Activity Pattern Averaged Across Days (na removed)"
    )
```

### What is the interval that contains the greatest average steps?
```{r ordered interval mean}
ordered.total.interval.steps <- total.interval.steps[order(-total.interval.steps$mean.steps), ]
head(ordered.total.interval.steps, 1)
```
## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
